---
title: "s06 Exercises: Pivoting and tidy data"
output: 
  html_document:
    keep_md: true
    theme: paper
---

```{r, warning = FALSE, message = FALSE}
library(tidyverse)
lotr  <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/lotr_tidy.csv")
guest <- read_csv("https://raw.githubusercontent.com/USF-Psych-DataSci/Classroom/master/data/wedding/attend.csv")
email <- read_csv("https://raw.githubusercontent.com/USF-Psych-DataSci/Classroom/master/data/wedding/emails.csv")
```

<!---The following chunk allows errors when knitting--->

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE, warning = FALSE)
```

## In class

```{r}
datasets::HairEyeColor

# NOT PRINTING BECAUSE THERE IS NO HAIREYE DATASET 
haireye_untidy <- haireye %>% 
  mutate(eye = str_c(eye, "_eyed")) %>% 
  pivot_wider(id_cols = hair, names_from = eye, values_from = n)
haireye_untidy

haireye_untidy %>% 
  pivot_longer(cols      = c(Blue_eyed, Brown_eyed, Green_eyed, Hazel_eyed), #making a new column that has some values in it. What names are the columns we want
               names_to  = "eye", #what name do we want to give that column that names the variables
               values_to = "n") #what values you want to give the column that has all of the values

haireye_untidy %>% 
  pivot_longer(cols      = Blue_eyed:Hazel_eyed, #this might be a problem later on because there could be new variables that are missed, values were switched and some are missing, there is no easy way to see what values are actually in your data.
               names_to  = "eye",
               values_to = "n")

# THIS OPTION SEEMS TO BE THE EASIEST TO USE BECAUSE IT IS CLEAR WHAT IS BEING CALLED
haireye_untidy %>% 
  pivot_longer(cols      = contains("eyed"), #this will pull in any variable name that contains whatever is specified here (eye)
               names_to  = "eye", # store all the column names that match in a column called
               values_to = "n") #store all the values in a column called

## PIVOT WIDER
haireye %>% 
  pivot_wider(id_cols     = hair,  # what are the ID columns
              names_from  = eye,  # what column are we getting the new column names from
              values_from = n)   # what column are we getting the values from

## FROM MULTIVARIATE
iris2 %>% 
  pivot_longer(cols      = c(-species, -id), # taking all the columns except these ones
               names_to  = c("part", "dimension"), #we now have a vector of names so it will create 2 new columns
               names_sep = "_",  # this is taking the information and separating the column names into do and splitting based on where the _ is. Everything before _ goes to part and everything after _ goes to dimension
               values_to = "measurement") # name column with the values

iris2 %>% 
  pivot_longer(cols      = c(-id, -species), 
               names_to  = c("part", ".value"), # the .value will say leave the part after _ as a separate column and leave the other part as the values for the "part" column
               names_sep = "_") #separate column names using the _ character

iris2_longest %>% 
  pivot_wider(id_cols     = c(id, species), #same as before
              names_from  = c(part, dimension), # making new columns that take part of the name from the part column and part of the name from the dimension column and using the _ to put them together for the new column names
              names_sep   = "_", 
              values_from = measurement)
```


## Exercise 1: Univariate Pivoting

Consider the Lord of the Rings data:

```{r}
lotr
```

1. Would you say this data is in tidy format?

This data is tidy if you are interested in seeing the number of words spoken by each race split by gender for each movie. If you are interested in the number of words by races or genders, then this is not a tidy dataset and it would need to be re-organized
   <!-- Describe why or why not in this space. -->

2. Widen the data so that we see the words spoken by each race, by putting race as its own column.

```{r}
(lotr_wide <- lotr %>% 
  pivot_wider(id_cols = c(-Race, -Words), 
              names_from = Race, 
              values_from = Words))
```

3. Re-lengthen the wide LOTR data from Question 2 above.

```{r}
lotr_wide %>% 
  pivot_longer(cols = c(Elf, Hobbit, Man), 
               names_to  = "Race", 
               values_to = "Words")
```


## Exercise 2: Multivariate Pivoting

Congratulations, you're getting married! In addition to the wedding, you've 
decided to hold two other events: a day-of brunch and a day-before round of 
golf. You've made a guestlist of attendance so far, along with food preference 
for the food events (wedding and brunch).

```{r}
guest %>% 
  DT::datatable(rownames = FALSE)
```

1. Put "meal" and "attendance" as their own columns, with the events living in a new column.

```{r}
(guest_long <- guest %>% 
  pivot_longer(cols      = c(meal_wedding, meal_brunch, attendance_wedding, attendance_brunch, attendance_golf), 
               names_to  = c(".value", "event"),
               names_sep = "_"))
```

2. Use `tidyr::separate()` to split the name into two columns: "first" and 
"last". Then, re-unite them with `tidyr::unite()`.

```{r}
guest_long %>% 
  separate(name, into = c("first", "last"))  %>% 
   unite(col = "name", c(first, last), sep = " ")
```

3. Which parties still have a "PENDING" status for all members and all events?

```{r}
guest_long %>% 
  group_by(party) %>% 
  summarize(all_pending = all(attendance == "PENDING"))
```

4. Which parties still have a "PENDING" status for all members for the wedding?

```{r}
guest_long %>% 
  group_by(party) %>% 
  filter(event == "wedding") %>% 
  summarize(pending_wedding = all(attendance == "PENDING"))
```


5. Put the data back to the way it was.

```{r}
guest_long %>% 
  pivot_wider(id_cols     = c(party, name), 
              names_from  = event, 
              names_sep   = "_", 
              values_from = c(meal, attendance)) %>% 
  select(-meal_golf)
```

6. You also have a list of emails for each party, in this worksheet under the 
   object `email`. Change this so that each person gets their own row. 
   Use `tidyr::separate_rows()`

```{r}
email %>% 
  separate_rows(guest, sep = ",")
```


## Exercise 3: Making tibbles

1. Create a tibble that has the following columns:

- A `label` column with `"Sample A"` in its entries.
- 100 random observations drawn from the N(0,1) distribution in the column `x`
  - "N" means the normal distribution. "(0, 1)" means mean = 0, sd = 1.
  - Use `rnorm()`
- `y` calculated as the `x` values + N(0,1) error. 

```{r}
n <- 100
FILL_THIS_IN(label = FILL_THIS_IN,
             FILL_THIS_IN = rnorm(n),
             FILL_THIS_IN)
```


2. Generate a Normal sample of size 100 for each combination of the following 
means (`mu`) and standard deviations (`sd`).

```{r}
n <- 100
mu <- c(-5, 0, 5)
sd <- c(1, 3, 10)
FILL_THIS_IN(mu = mu, sd = sd) %>% 
  group_by_all() %>% 
  mutate(z = list(rnorm(n, mu, sd))) %>% 
  FILL_THIS_IN
```


3. Fix the `experiment` tibble below (originally defined in the documentation 
of the `tidyr::expand()` function) so that all three repeats are displayed for 
each person, and the measurements are kept. Some code is given, but it doesn't
quite work. It needs a few adjustments. What are they?

```{r}
experiment <- tibble(
  name = rep(c("Alex", "Robert", "Sam"), c(3, 2, 1)),
  trt  = rep(c("a", "b", "a"), c(3, 2, 1)),
  rep = c(1, 2, 3, 1, 2, 1),
  measurement_1 = runif(6),
  measurement_2 = runif(6)
)
experiment %>% expand(name, trt, rep)
```


