---
title: "cm008 Exercises: Fix the Plots"
output: 
  html_document:
    keep_md: true
    theme: paper
---

In this worksheet, we'll be looking at some erroneous plots and fixing them. 

I think you might not have these two packages installed:

```
install.packages("ggridges")
install.packages("scales")
```

The exercsies below include some data wrangling function. It's okay if you
aren't familiar with them all yet! We will get into a lot of them over the
next few weeks, but see if you can figure out what they are doing as you go.

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(gapminder)
library(ggridges)
library(scales)
```

<!-- The following chunk allows errors when knitting -->

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE, warning = FALSE)
```


## Exercise 1: Overlapping Points

Fix the overlapping data points problem in the following plot by adding an `alpha`
or `size` argument (attribution: ["R for data science"](https://r4ds.had.co.nz/data-visualisation.html)).

```{r, fig.width = 5, fig.height = 2}
ggplot(mpg, aes(cty, hwy)) + 
  geom_point(alpha = 0.1)
```


## Exercise 2: Line for each Country

Fix this plot so that it shows life expectancy over time with a separate line
_for each country_. 

Notice that we tried to use `group_by()`. We will cover that next week. But 
also notice that `ggplot2` ignores the grouping of a tibble!

```{r, fig.width = 5, fig.height = 2}
gapminder %>% 
  group_by(country) %>% 
  ggplot(aes(year, lifeExp)) +
  geom_line(aes(group = country))
```


## Exercise 3: More gdpPercap vs lifeExp

### 3(a) Facets

- Change the x-axis text to be in "comma format" with `scales::comma_format()`.
- Separate each continent into sub-panels.

```{r, fig.width = 5, fig.height = 2}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
    geom_point(alpha = 0.2) +
  scale_x_log10(labels = scales::comma_format())+
  facet_wrap(vars(continent))
```

### 3(b) Bubble Plot

- Put the plots in one row, and free up the axes (let them have different scales).
- Make a bubble plot by making the size of the points proportional to population. 
  - Try adding a `scale_size_area()` layer too (could also try `scale_radius()` 
    but that is not optimal for perception).
- Use `shape=21` to distinguish between `fill` (interior) and `color` (exterior). 
My original version
```{r, fig.width = 5, fig.height = 2}
gapminder %>% 
  filter(continent != "Oceania") %>% 
  ggplot(aes(gdpPercap, lifeExp)) +
  facet_wrap(~ continent) + # this is breaking it into different panels
  geom_point(alpha = 0.2) +
  scale_x_log10(labels = scales::comma_format()) +
  scale_size_area(aes(shape = 21))
```

Corrected version using posted key:
```{r, fig.width = 5, fig.height = 2}
gapminder %>% 
  filter(continent != "Oceania") %>% 
  ggplot(aes(gdpPercap, lifeExp, size = pop, fill = continent)) + #changes the size and fill of the dots based on the population and continent
  facet_wrap(~ continent, nrow = 1) + # putting everything into a single row
  geom_point(alpha = 0.2, shape = 21) +
  scale_x_log10(labels = scales::comma_format()) +
  scale_size_area() +
  scale_fill_discrete(guide = FALSE)
```


A list of shapes can be found [at the bottom of the `scale_shape` documentation](https://ggplot2.tidyverse.org/reference/scale_shape.html).

### 3(c) Size "not working"

Instead of alpha transparency, suppose you're wanting to fix the overplotting issue by plotting small points. Why is this not working? Fix it.

It was not working because the size value should not be included within the aes() function.
```{r, fig.width = 5, fig.height = 2}
ggplot(gapminder) +
  geom_point(size = 0.1, aes(gdpPercap, lifeExp)) +
  scale_x_log10(labels = scales::dollar_format())
```


## Exercise 4: Walking caribou

The following mock data set marks the (x,y) position of a caribou at four time points. 

- Fix the plot below so that it shows the path of the caribou. 
- Add an arrow with `arrow = arrow()`.
- Add the `time` label with `geom_text()`.

```{r, fig.width = 5, fig.height = 2}
tribble(
  ~time, ~x, ~y,
  1, 0.3, 0.3,
  2, 0.8, 0.7,
  3, 0.5, 0.9,
  4, 0.4, 0.5
) %>% 
  ggplot(aes(x, y)) + 
  #geom_line() + #removing this as a correction from the answer key
  geom_path(arrow = arrow()) +
  geom_text(aes(label = c("Time 1", "Time 2", "Time 3", "Time 4")))
```

## Exercise 5: Life expectancies in Africa

### 5(a) Unhiding the data

Fix the plot so that you can actually see the data points. 

There is also the problem of overlapping text in the x-axis labels. How could we solve that?
I abbreviated and rotated the x-axis labels. Code to fix the x-axis labels from [here](https://stackoverflow.com/questions/1828742/rotating-axis-labels-in-r)

```{r, fig.width = 5, fig.height = 2}
gapminder %>% 
  filter(continent == "Americas") %>% 
  ggplot(aes(country, lifeExp)) + 
  geom_boxplot() +
  geom_point(alpha = 0.2) +
  scale_x_discrete(labels = abbreviate) +
  theme(axis.text.x = element_text(angle=45, hjust=1)) #from stackoverflow
```

### 5(b) Ridgeplots

We're starting with the same plot as above, but instead of the points + boxplot, try a ridge plot instead using `ggridges::geom_density_ridges()`, and adjust the `bandwidth`.

```{r, fig.width = 5, fig.height = 2}
gapminder %>% 
  filter(continent == "Americas") %>% 
  ggplot(aes(lifeExp, country)) + 
  ggridges::geom_density_ridges(aes(scale = 2)) #allows for the adjustment of the hills for the lines (whether they are taller or flatter)
```

## Exercise 6: Bar plot madness

### 6(a) Colour and stacking madness

- Change the following plot so that it shows _proportion_ on the y-axis, not count.
- Change the x-axis so that it doesn't appear to be continuous.
  - Hint: Transform the variable!
- Also change the colors of the bar fills, as well as the lines.
- Put the bars for transmission side-by-side with their own colour.
- Capitalize the legend title.

Notes: I was able to get the proportions on the y and the variables discrete but they removed all of the x-labels and I cannot get them to return. Not Sure how to get the Transmission capitalized at this time.

```{r, fig.width = 5, fig.height = 2}
mtcars %>% 
  mutate(transmission = if_else(am == 0, "automatic", "manual")) %>%
  ggplot(aes(cyl)) +
  geom_bar(position = "dodge", aes(color = transmission, y = ..prop.., fill = transmission)) +
  scale_x_continuous(breaks = c(4, 6, 8)) #this was provided from the key and replaced the old code.
  
  # labs(ledgend = "Transmission")
```

### 6(b) Bar heights already calculated

Here's the number of people having a certain hair colour from a sample of 592 people:

```{r}
(hair <- as_tibble(HairEyeColor) %>% 
  count(Hair, wt = n))
```

Fix the following bar plot so that it shows these counts.

```{r}
ggplot(hair, aes(Hair, n)) +
  geom_col() +
  ylab("Count")
```

## Exercise 7: Tiling

Here's the number of people having a certain hair and eye colour from a sample of 592 people:

```{r}
(hair_eye <- as_tibble(HairEyeColor) %>% 
  count(Hair, Eye, wt = n))
```

Fix the following plot so that it shows a filled-in square for each combination. 
_Hint:_ What's the title of this exercise?

```{r}
ggplot(hair_eye, aes(Hair, Eye)) +
  geom_point(aes(colour = n)) +
  geom_tile(aes(fill = n))
```

By the way, `geom_count()` is like `geom_bar()`: it counts the number of overlapping points.


## Additional take-home practice

If you'd like some practice, give these exercises a try

__Exercise 1__: Make a plot of `year` (x) vs `lifeExp` (y), with points coloured by continent. Then, to that same plot, fit a straight regression line to each continent, without the error bars. If you can, try piping the data frame into the `ggplot()` function.
Help for regression line found [here](https://stackoverflow.com/questions/15633714/adding-a-regression-line-on-a-ggplot)

```{r}
ggplot(gapminder, aes(year, lifeExp, color = continent)) + #adding color here as suggested in the key and removed from the geom_point aes(). If it is only in the point line then only the points will be colored and the lines will not be- there was only one line.
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE)
```


__Exercise 2__: Repeat Exercise 1, but switch the _regression line_ and _geom\_point_ layers. How is this plot different from that of Exercise 1?

Initially the graph looks the same but looking closer shows that since the coding order was reversed, the line is actually behind the points in this graph whereas in the first graph, the line appeared overtop of the points. The standard error can be added to the line by changing the se = FALSE to se = TRUE.

```{r}
ggplot(gapminder, aes(year, lifeExp, color = continent)) + 
 geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  geom_point(alpha = 0.2) 
#also moved color to the ggplot line
```

__Exercise 3__: Omit the `geom_point()` layer from either of the above two plots (it doesn't matter which). Does the line still show up, even though the data aren't shown? Why or why not?

Removing the points from the graph did not remove the line from the graph because they are using two separate functions. The geom_point() function creates a graph that uses only points to represent the data. The geom_smooth() function creates a line for the data, in this case a regression line but it can also be log or a unique formula if specified. Removing the grom_point() function just removed one of the layers of the graph.

```{r}
ggplot(gapminder, aes(year, lifeExp, color = continent)) + #added color to ggplot aes()
  #geom_point(alpha = 0.2, aes(color = continent)) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE)
```

__Exercise 4__: Make a plot of `year` (x) vs `lifeExp` (y), facetted by continent. Then, fit a smoother through the data for each continent, without the error bars. Choose a span that you feel is appropriate.

```{r, fig.width = 8, fig.height = 2}
ggplot(gapminder, aes(year, lifeExp)) +
    geom_line(alpha = 0.5, aes(group = country, color = continent)) +
  facet_wrap(vars(continent), nrow = 1) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  theme(axis.text.x = element_text(angle=45, hjust=1)) #from stackoverflow (see exercise 5a)
```
Some changes to the code made based on the key, like making them into one row

__Exercise 5__: Plot the population over time (year) using lines, so that each country has its own line. Colour by `gdpPercap`. Add alpha transparency to your liking. 

```{r, fig.width = 5, fig.height = 2}
  ggplot(gapminder, aes(year, pop)) +
  geom_line(aes(alpha = .5, group = country, color = gdpPercap)) +
  scale_y_continuous(labels = scales::comma_format())
```
 
Another version (**Preferred version**) using the log 10 for the y axis to be able to see the lines better:
```{r, fig.width = 5, fig.height = 2}
  ggplot(gapminder, aes(year, pop)) +
  geom_line(aes(alpha = .5, group = country, color = gdpPercap)) +
  scale_y_log10(labels = scales::comma_format())
```


__Exercise 6__: Add points to the plot in Exercise 5.

First version of 5
```{r, fig.width = 5, fig.height = 2}
  ggplot(gapminder, aes(year, pop)) +
  geom_line(aes(alpha = .5, group = country, color = gdpPercap)) +
  scale_y_continuous(labels = scales::comma_format()) +
  geom_point()
```

Second version (**Preferred version**) of 5
```{r, fig.width = 5, fig.height = 2}
  ggplot(gapminder, aes(year, pop)) +
  geom_line(aes(alpha = .5, group = country, color = gdpPercap)) +
  scale_y_log10(labels = scales::comma_format()) +
  geom_point(aes(alpha = .5))
```